---
title: 'Clustering with multiple constraints and attributes: Simulated example'
author: "Tero Lähderanta"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, load the necessary libraries. In addition to the `rpack` package, we use `tidyverse` and `dplyr` for sample data manipulation and `ggplot2` for plotting.

```{r message=FALSE}
#install.packages("lcmix", repos=c("http://R-Forge.R-project.org",
#                                   "http://cran.at.r-project.org"),dependencies=TRUE)
library(rpack)
library(tidyverse)
library(LaplacesDemon)
library(Matrix)
library(plotly)
library(Gmedian)
library(lcmix) # Added by Markku. Package can be found at R-Forge
library(parallel) # For parallel computing. rpack uses function "detectCores".
library(flexmix)
library(patchwork)
library(purrr)
library(broman)
#devtools::install_github("jamesotto852/ggdensity")
library(ggdensity)
library(targets)
library(MASS)
library(conflicted)
library(scales)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

These external functions generate data from gamma distribution. Weights are determined either by 1) chance 2 ) based on the distance from the distribution mean: the greater the Euclidean distance between the point and the population mean is, the larger the weight is. `simulate_unif_grid` is a sub-function of the main function `simulate_gamma_mixture`.

```{r}
source("CodeCollection/simulate_gamma_mixture.R")
source("CodeCollection/simulate_unif_grid.R")
source("CodeCollection/utility_functions.R")
source("functions_rpack.R")
source("functions_summary.R")

# Get the colors for plotting
c_col <- get_cols() 
```

# Evaluation with simulated data

In this section we introduce the properties of PACK extensions with a controlled setup. First we generate multiple random data sets, where the original clustering is known and consequently we can analyze the goodness of the algorithm with the selected parameters. 

## Simulated data

Data points are simulated from a gamma mixture distribution with varying shape and scale parameters and the weigths of the data points are simulated from a uniform distribution, such that the weigths are between 1 and 100. 

Let's set up some data to be clustered.

-   To ease cluster overlap, clusters are placed on a grid

-   Outliers are sampled uniformly on the cluster grid. Some simulated clusters have heavy weight points at edge of clusters (50/50 chance)\*\*

-   We study the scalability of the algorithm by simulating different sizes of data sets

-   Different non-spatial attributes are added to the analysis


Next, we create non-spatial attributes by dividing data points to four sections as shown below.


```{r warning=FALSE}

dat <- create_data_example(seed = 329689)

# Plot the example data 
plot_division_example(dat)
# ggsave(filename = "images/plot_division_example.png",
#        width = 5,
#        height = 5)
```

Visualize the distribution of non-spatial attributes in each section.
```{r}
# Parameters for the simulated distribution of non-spatial attributes
params <- tribble(
    ~group, ~mu, ~Sigma,
    1, c(1, 5, 7), diag(c(1,2,2)),
    2, c(4, 5, 6), diag(c(1,0.5,1)),
    3, c(8, 4, 1), diag(c(2,2,2)),
    4, c(5, 7, 3), diag(c(1,1,2)),
  )

dat$Y |> 
  pivot_longer(cols = c("par1", "par2", "par3"),
               names_to = "non_spat_par") |>
  mutate(non_spat_group = paste("Section", non_spat_group)) |> 
  ggplot()+
  geom_boxplot(aes(y = value, 
                   #x = non_spat_group,
                   color = non_spat_par)) +
  facet_grid(cols = vars(non_spat_group)) +
  guides(color=guide_legend(title="Non-spatial attribute:")) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom")
# ggsave(filename = "images/plot_nonspat_example.png",
#        width = 6,
#        height = 5)
```

```{r}
colpal <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
plot_point_gradient(dat$Y,
                    "par1",
                    colpal[1]) /
  plot_point_gradient(dat$Y,
                      "par2",
                      colpal[2]) /
  plot_point_gradient(dat$Y,
                      "par3",
                      colpal[3])

# ggsave(filename = "images/plot_nonspat_distribution1.png",
#        width = 4,
#        height = 9)
```

```{r}
dat$Y |> 
  mutate(r_par1 = rescale(par1, c(0,1)),
         g_par2 = rescale(par2, c(0,1)),
         b_par3 = rescale(par3, c(0,1))) |> 
  ggplot() +
  geom_point(aes(x = x, y = y, color = rgb(r_par1, g_par2, b_par3), size = w)) +
  scale_size(range = c(2, 6)) +
  guides(size = "none", color = "none") +
    labs(x = "x", y = "y") +
    theme(
      legend.position = "right",
      axis.text.x = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_blank(),
      axis.ticks = ggplot2::element_blank()
    )+
    coord_fixed(ratio = 1)

# ggsave(filename = "images/plot_nonspat_distribution2.png",
#        width = 5,
#        height = 4)

```

Let's then simulate $N$ data sets from gamma mixture with different data points and non-spatial attributes. 

```{r}
# Simulate 4 data sets from gamma mixture
tar_load(list_dat100)
```

Visualize the data sets:

```{r warning=FALSE}
# Make plot from each list element
plots_dat100 <- list_dat100 |> 
  map(plot_nonspat_attributes) 

plot_pw <- 
plots_dat100[[1]] + 
  plots_dat100[[2]] + 
  plots_dat100[[3]] + 
  plots_dat100[[4]] + 
  patchwork::plot_layout(guides = "collect")

plot_pw  
```

To generate non-spatial attributes, we simulate one value from 3D multinormal distribution. Each non-spatial group has different parameters for the normal distribution. 

```{r}
plot_pars <- list_dat100[[1]]$Y |> 
    ggplot() +
    geom_point(aes(x = x, 
                   y = y, 
                   #size = w, 
                   color = orig_group,
                   fill = orig_group,
                   shape = is_outlier,
                   label = par1,
                   label2 = par2,
                   label3 = par3), 
               size = 3,
               stroke = 1.5) +
    geom_vline(xintercept = list_dat100[[1]]$div[1]) + 
    geom_hline(yintercept = list_dat100[[1]]$div[2]) + 
    guides(size = "none", fill = "none") +
    scale_color_discrete(name = "Original clusters") +
    scale_shape_manual(values = c(21,4)) +
    guides(colour = guide_legend(override.aes = list(size=5)),
           shape = "none")

ggplotly(plot_pars, tooltip = c("par1", "par2", "par3"))
```


## Clustering

```{r}
tar_load(clust1)
tar_load(clust2)
tar_load(clust3)
tar_load(clust4)

##clust_dat4[[1]]
plot_clust1 <- plot_clust_list(
  list_dat100,
  clust1
)

plot_clust1

plot_clust2 <- plot_clust_list(
  list_dat100,
  clust2
)

plot_clust2

plot_clust3 <- plot_clust_list(
  list_dat100,
  clust3
)

plot_clust3

plot_clust4 <- plot_clust_list(
  list_dat100,
  clust4
)

plot_clust4
```

```{r, eval = FALSE}
clust_names <- paste("lambda=", seq(0.1, 1, by = 0.1), sep="")

clust1_proximity_summary <-
  proximity_summary(clust_list = clust1[[1]][-1],
                    names = clust_names,
                    dat = list_dat100[[1]]$Y)

clust1_proximity_summary

clust1_duration_summary = duration_sd_summary(
    clust_list = clust1[[1]][-1],
    names = clust_names,
    dat = list_dat100[[1]]$Y
  )

clust1_duration_summary

```

