---
title: 'Clustering with multiple constraints and attributes: Simulated example'
author: "Tero Lähderanta"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, load the necessary libraries. 

```{r message=FALSE}
#install.packages("lcmix", repos=c("http://R-Forge.R-project.org",
#                                   "http://cran.at.r-project.org"),dependencies=TRUE)
library(rpack)
library(tidyverse)
library(LaplacesDemon)
library(Matrix)
#library(plotly)
library(Gmedian)
library(lcmix) # Added by Markku. Package can be found at R-Forge
library(parallel) # For parallel computing. rpack uses function "detectCores".
#library(flexmix)
library(patchwork)
library(purrr)
library(broman)
#devtools::install_github("jamesotto852/ggdensity")
#library(ggdensity)
library(targets)
library(MASS)
library(scales)
library(progress)
#devtools::install_github("cmartin/ggConvexHull")
library(ggConvexHull)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")

# Load external functions
source("CodeCollection/simulate_gamma_mixture.R")
source("CodeCollection/simulate_unif_grid.R")
source("CodeCollection/utility_functions.R")
source("functions_rpack.R")
source("functions_summary.R")

# Get the colors for plotting
colpal <- get_cols() 

# Set ggplot theme
theme_set(theme_bw())

# Load targets
tar_load(c("list_dat100", "clust1", "clust2", "clust3", "clust4"))
```

These external functions generate data from gamma distribution. Weights are determined either by 1) chance 2 ) based on the distance from the distribution mean: the greater the Euclidean distance between the point and the population mean is, the larger the weight is. `simulate_unif_grid` is a sub-function of the main function `simulate_gamma_mixture`.

# Evaluation with simulated data

In this section we introduce the properties of PACK extensions with a controlled setup. First we generate multiple random data sets, where the original clustering is known and consequently we can analyze the goodness of the algorithm with the selected parameters. 

## Simulated data

Data points are simulated from a gamma mixture distribution with varying shape and scale parameters and the weigths of the data points are simulated from a uniform distribution, such that the weigths are between 1 and 100. 

Let's set up some data to be clustered.

-   To ease cluster overlap, clusters are placed on a grid

-   Outliers are sampled uniformly on the cluster grid. Some simulated clusters have heavy weight points at edge of clusters (50/50 chance)\*\*

-   We study the scalability of the algorithm by simulating different sizes of data sets

-   Different non-spatial attributes are added to the analysis

### Single example data set

Next, we create non-spatial attributes by dividing data points to four sections as shown with an example data set below.

```{r warning=FALSE}
# Crete example data with set seed
dat <- create_data_example(seed = 329689)

# Plot the example data 
plot_division_example(dat)
# ggsave(filename = "images/plot_division_example.png",
#        width = 5,
#        height = 5)
```

Visualize the distribution of non-spatial attributes in each section.

```{r}
# Parameters for the simulated distribution of non-spatial attributes
params <- tribble(
    ~group, ~mu, ~Sigma,
    1, c(1, 5, 7), diag(c(1,2,2)),
    2, c(4, 5, 6), diag(c(1,0.5,1)),
    3, c(8, 4, 1), diag(c(2,2,2)),
    4, c(5, 7, 3), diag(c(1,1,2)),
  )

# Plot the boxplots of each non-spatial parameter in each region
dat$Y |> 
  pivot_longer(cols = c("par1", "par2", "par3"),
               names_to = "non_spat_par") |>
  mutate(non_spat_group = paste("Region", non_spat_group)) |> 
  ggplot()+
  geom_boxplot(aes(y = value, 
                   #x = non_spat_group,
                   color = non_spat_par)) +
  facet_grid(cols = vars(non_spat_group)) +
  guides(color=guide_legend(title="Non-spatial attribute:")) +
  scale_color_manual(labels = c("Red", "Green", "Blue"), values = 
                       c("#F8766D","#7CAE00", "#00BFC4")) + 
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom")
ggsave(filename = "images/plot_nonspat_example.png",
       width = 6,
       height = 3)
```

```{r}
# Plot the non-spatial parameter distribution in three plots
plot_point_gradient(dat$Y,
                    "par1",
                    colpal[1]) /
  plot_point_gradient(dat$Y,
                      "par2",
                      colpal[2]) /
  plot_point_gradient(dat$Y,
                      "par3",
                      colpal[3])

# ggsave(filename   = "images/plot_nonspat_distribution1.png",
#        width = 4,
#        height = 9)
```

```{r}
# Plot RGB point plot for the non-spatial attributes 
dat$Y |> 
  mutate(r_par1 = rescale(par1, c(0,255)) |> round(),
         g_par2 = rescale(par2, c(0,255)) |> round(),
         b_par3 = rescale(par3, c(0,255)) |> round()) |> 
  mutate(color = rgb(red = r_par1, green = g_par2, blue = b_par3, 
                     maxColorValue = 255)) |> 
  ggplot() +
  geom_point(aes(x = x, y = y, 
                 size = w, 
                 color = color)) +
  scale_color_identity() +
  scale_size(range = c(2, 6)) +
  guides(size = "none", color = "none") +
  labs(x = "x", y = "y") +
  theme_bw() +
  coord_fixed(ratio = 1)
# ggsave(filename = "images/plot_nonspat_distribution2.png",
#        width = 5,
#        height = 5)

```

Next, cluster the example data set into 10 clusters with rpack. We first set the clustering parameters.

```{r}
# Set range and lambda_o
mean_dat <- round(sum(dat$Y |> pull(w) / 10))
range <-  c((mean_dat - 10*100), (mean_dat + 10*100))
lambda_par <- seq(0.1, 1, 0.1)
```

Cluster with varying $\lambda_o$ and $\lambda_d$.

```{r}
N <- 50

# clust_dat <-
#   clust_with_params(
#     coords = dat$Y |> dplyr::select(x, y),
#     params = dat$Y |> dplyr::select(par1, par2, par3),
#     weights = dat$Y |> pull(w),
#     lambda_par = lambda_par, 
#     lambda = NULL,
#     k = 10,
#     N = N,
#     range = range
#   )
# 
# clust_dat2 <-
#   clust_with_params(
#     coords = dat$Y |> dplyr::select(x, y),
#     params = dat$Y |> dplyr::select(par1, par2, par3),
#     weights = dat$Y |> pull(w),
#     lambda_par = lambda_par, 
#     lambda = 0.3,
#     k = 10,
#     N = N,
#     range = range
#   )
# 
# clust_dat3 <-
#   clust_with_params(
#     coords = dat$Y |> dplyr::select(x, y),
#     params = dat$Y |> dplyr::select(par1, par2, par3),
#     weights = dat$Y |> pull(w),
#     lambda_par = lambda_par, 
#     lambda = 0.25,
#     k = 10,
#     N = N,
#     range = range
#   )
# 
# clust_dat4 <-
#   clust_with_params(
#     coords = dat$Y |> dplyr::select(x, y),
#     params = dat$Y |> dplyr::select(par1, par2, par3),
#     weights = dat$Y |> pull(w),
#     lambda_par = lambda_par, 
#     lambda = 0.2,
#     k = 10,
#     N = N,
#     range = range
#   )
# clust_dat5 <-
#   clust_with_params(
#     coords = dat$Y |> dplyr::select(x, y),
#     params = dat$Y |> dplyr::select(par1, par2, par3),
#     weights = dat$Y |> pull(w),
#     lambda_par = lambda_par, 
#     lambda = 0.15,
#     k = 10,
#     N = N,
#     range = range
#   )
# 
# clust_dat6 <-
#   clust_with_params(
#     coords = dat$Y |> dplyr::select(x, y),
#     params = dat$Y |> dplyr::select(par1, par2, par3),
#     weights = dat$Y |> pull(w),
#     lambda_par = lambda_par, 
#     lambda = 0.1,
#     k = 10,
#     N = N,
#     range = range
#   )


# save(clust_dat, file= "temp_files\\clust_dat.Rdata")
# save(clust_dat2, file= "temp_files\\clust_dat2.Rdata")
# save(clust_dat3, file= "temp_files\\clust_dat3.Rdata")
# save(clust_dat4, file= "temp_files\\clust_dat4.Rdata")
# save(clust_dat5, file= "temp_files\\clust_dat5.Rdata")
# save(clust_dat6, file= "temp_files\\clust_dat6.Rdata")
load("~/R-projects/FMM_artificialdata/temp_files/clust_dat.Rdata")
load("~/R-projects/FMM_artificialdata/temp_files/clust_dat2.Rdata")
load("~/R-projects/FMM_artificialdata/temp_files/clust_dat3.Rdata")
load("~/R-projects/FMM_artificialdata/temp_files/clust_dat4.Rdata")
load("~/R-projects/FMM_artificialdata/temp_files/clust_dat5.Rdata")
load("~/R-projects/FMM_artificialdata/temp_files/clust_dat6.Rdata")

```

Plot all the summary plots.

```{r}
clust_names <- paste("lambda=", seq(0.1, 1, by = 0.1), sep="")

full_summary_dat <- full_summary(clust_dat, clust_names)
full_summary_dat2 <- full_summary(clust_dat2, clust_names)
full_summary_dat3 <- full_summary(clust_dat3, clust_names)
full_summary_dat4 <- full_summary(clust_dat4, clust_names)
full_summary_dat5 <- full_summary(clust_dat5, clust_names)
full_summary_dat6 <- full_summary(clust_dat6, clust_names)



plot_summary_lines(full_summary_dat, scale_coeff = 0.118) +
  plot_summary_lines(full_summary_dat2, scale_coeff = 0.118) +
  plot_summary_lines(full_summary_dat3, scale_coeff = 0.118) +
  plot_summary_lines(full_summary_dat4, scale_coeff = 0.118) +
  plot_summary_lines(full_summary_dat5, scale_coeff = 0.118) +
  plot_summary_lines(full_summary_dat6, scale_coeff = 0.118) +
plot_layout(guides = "collect", ncol = 2 ) & theme(legend.position = 'bottom')
```

```{r}
plot_summary_lines(full_summary_dat |> 
                     mutate(mean_sd_par3 = mean_sd_par3^2), scale_coeff = 0.118)
```


```{r}
(plot_summary_lines(full_summary_dat, 0.118) + ggtitle(expression(lambda[o]*" = 1"~(No~outliers)))) /   
  (plot_summary_lines(full_summary_dat4, 0.118) + ggtitle(expression(lambda[o]*" = 0.2"))) /
  (plot_summary_lines(full_summary_dat6, 0.118) + ggtitle(expression(lambda[o]*" = 0.1"))) +
  plot_layout(guides = "collect", ncol = 1) & 
  theme(legend.position = 'bottom') & 
  guides(color = guide_legend(nrow = 2, byrow = TRUE))

ggsave(filename = "images/singleclust_summary.png",
       width = 6,
       height = 9)
```



Plot some of the clusterings.

```{r}
(plot_cl_facet(clust_dat[-1], dat, c(2,6,10)) + ggtitle(expression(lambda[o]*" = 1"~(No~outliers)))) /  
(plot_cl_facet(clust_dat4[-1], dat, c(2,6,10))+ ggtitle(expression(lambda[o]*" = 0.2"))) /
(plot_cl_facet(clust_dat6[-1], dat, c(2,6,10))+ ggtitle(expression(lambda[o]*" = 0.1"))) +
  plot_layout(guides = "collect", ncol = 1)

ggsave(filename = "images/singleclust_plots.png",
       width = 6,
       height = 8)
```


### Summary of 100 data sets

Let's then simulate $N$ data sets from gamma mixture with different data points and non-spatial attributes.

To generate non-spatial attributes, we simulate one value from 3D multinormal distribution. Each non-spatial group has different parameters for the normal distribution. 


## Clustering


```{r}
clust_names <- paste("lambda=", seq(0.1, 1, by = 0.1), sep="")

proximity_summary_table1 <- proximity_summary_list(clust_list = clust1,
                       dat_list = list_dat100,
                       clust_names = clust_names
                       )

duration_summary_table1 <- duration_summary_list(clust_list = clust1,
                       dat_list = list_dat100,
                       clust_names = clust_names
                       )

proximity_summary_table1 <- proximity_summary_table1 |>
  select(name, Mean) |> 
  group_by(name) |> 
  summarise(Mean = mean(Mean))

duration_summary_table1 <- duration_summary_table1 |>
  select(name, Mean) |> 
  group_by(name) |> 
  summarise(Mean = mean(Mean))

# Combined table
full_summary1 <- duration_summary_table1 |> 
  mutate(par = str_sub(name, 1,4), .after = 1) |> 
  mutate(name = str_sub(name, 6,-1) |> str_replace("lambda", "lambda=")) |> 
  pivot_wider(values_from = Mean, names_from = par, names_prefix = "mean_sd_") |> 
  full_join(proximity_summary_table1, by = "name")

```

```{r}
clust_names <- paste("lambda=", seq(0.1, 1, by = 0.1), sep="")

proximity_summary_table2 <- proximity_summary_list(clust_list = clust2,
                       dat_list = list_dat100,
                       clust_names = clust_names
                       )

duration_summary_table2 <- duration_summary_list(clust_list = clust2,
                       dat_list = list_dat100,
                       clust_names = clust_names
                       )

proximity_summary_table2 <- proximity_summary_table2 |>
  select(name, Mean) |> 
  group_by(name) |> 
  summarise(Mean = mean(Mean))

duration_summary_table2 <- duration_summary_table2 |>
  select(name, Mean) |> 
  group_by(name) |> 
  summarise(Mean = mean(Mean))

# Combined table
full_summary2 <- duration_summary_table2 |> 
  mutate(par = str_sub(name, 1,4), .after = 1) |> 
  mutate(name = str_sub(name, 6,-1) |> str_replace("lambda", "lambda=")) |> 
  pivot_wider(values_from = Mean, names_from = par, names_prefix = "mean_sd_") |> 
  full_join(proximity_summary_table2, by = "name")
```


```{r}
clust_names <- paste("lambda=", seq(0.1, 1, by = 0.1), sep="")

proximity_summary_table3 <- proximity_summary_list(clust_list = clust3,
                       dat_list = list_dat100,
                       clust_names = clust_names
                       )

duration_summary_table3 <- duration_summary_list(clust_list = clust3,
                       dat_list = list_dat100,
                       clust_names = clust_names
                       )

proximity_summary_table3 <- proximity_summary_table3 |>
  select(name, Mean) |> 
  group_by(name) |> 
  summarise(Mean = mean(Mean))

duration_summary_table3 <- duration_summary_table3 |>
  select(name, Mean) |> 
  group_by(name) |> 
  summarise(Mean = mean(Mean))

# Combined table
full_summary3 <- duration_summary_table3 |> 
  mutate(par = str_sub(name, 1,4), .after = 1) |> 
  mutate(name = str_sub(name, 6,-1) |> str_replace("lambda", "lambda=")) |> 
  pivot_wider(values_from = Mean, names_from = par, names_prefix = "mean_sd_") |> 
  full_join(proximity_summary_table3, by = "name")

```

Collect summary data into a single plot.

```{r}
(plot_summary_lines(full_summary1, 0.118) + ggtitle(expression(lambda[o]*" = 1"~(No~outliers)))) /
(plot_summary_lines(full_summary2, 0.118) + ggtitle(expression(lambda[o]*" = 0.1"))) /
(plot_summary_lines(full_summary3, 0.118) + ggtitle(expression(lambda[o]*" = 0.05"))) +
  patchwork::plot_layout(guides = "collect") & 
  theme(legend.position = 'bottom') & 
  guides(color = guide_legend(nrow = 2, byrow = TRUE))

ggsave(filename = "images/plot_clust_summary.png",
       width = 6,
       height = 9)
```

